Nextstrain: real-time tracking of virus evolution

### 2017-09-06 whitepaper.tex
Fix references

Viral pathogens pose an ongoing and ever present danger to human health on a worldwide scale, including epidemics such as the West-African Ebola epidemic and the ongoing Zika epidemic in the Americas, as well as ongoing disease spread such as seasonal influenza. 
The rapid evolution of these viruses allows inference of the evolution and epidemiology of the disease from genomic data. 
Such analysis are often done in isolation, and may lack the spatial or temporal context in which to best interpret the results, which impedes understanding \citep{Pybus:2013bj}.
Furthermore, the results of analysis are often not made available to the public or health bodies until publication, which may be too late to effect change in policy or understanding. 
We have developed nextstrain to visualise viral outbreaks in as close to real time as possible.
This platform consists of three components: the collection and updating of available sequences from a range of public and private sources, rapid and automated analysis pipelines and an interactive web-application by which to interrogate the results.

The regularly updated nature and rapidity of these analysis is crucial to the monitoring and understanding of pathogen epidemiology and evolution.  
Sequencing times and costs are continually dropping, with on-the-ground sequencing being employed during the recent West African Ebola and Zika epidemics \citep{Quick:2016kf,Faria105171}.
This rapidity of sequencing must be complemented with rapid methods by which to analyse and interpret results.

Nextstrain consists of three components:
``Fauna'' is a collection of python scripts to maintain a database representative of all publicly available sequences and related metadata, sourced from public repositories such as NCBI, gisaid and ViPR, as well as github repositories and other sources of genomic data. ``Augur'' is the bioinformatics pipeline, performing subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait reconstruction across the tree, including inference of the most likely transmission events.
This leverages the maximum likelihood phylodynamic analysis TreeTime \citep{Sagulenko:2017vm}, allowing a full analysis of the entire Ebola epidemic ($n=000$) in 0 hours. 
These scripts are designed in a manner to be readily co-opted to different pathogens.
Analysed data are then immediately available to interrogate through the visualisation software ``Auspice'', available at \href{www.nextstrain.org}{www.nextstrain.org}.
This approach is similar in concept to Nextflu \citep{neher2015nextflu} however extended and generalised to different viral pathogens.
There is a growing need for surveillance of non-influenza viruses \citep{FRCPath:2017fg}, and nextstrain is able to be extended to nearly all viral outbreaks with readily accessible genomic data.

@@

For many pathogens, the emergence and spread of gain-of-function variants is a grave concern.
China has experienced seasonal epidemics of influenza A/H7N9 over the past five years, however human to human transmission is thought to be rare, with the majority of human cases being spillover from poultry. 
The threat of mutations which facilitate human to human transmission is of extreme concern, as H7N9 has a mortality rate of around $30\%$ \citep{doi:10.1056/NEJMoa1304617}.
\textit{In vitro} experiments have identified three amino acid mutations in the hemagglutinin (HA) gene which confer a switch in specificity for human-type receptors by promoting binding to human trachea epithelial cells \citep{deVries:2017dj}. 
Nextstrain allows monitoring of such mutations, such that interventions and control measures may be implemented where appropriate.

\subsection*{A Model for Public Sharing of Data}
Nextstrain presents a single overview of both ongoing viral disease (seasonal \& avian influenza, zika) as well as retrospective analysis (dengue, ebola), all based upon the same underlying bioinformatics architecture.
As such, it is well positioned to respond to both the next outbreak and continued surveillance of current ones. 
Analysis of such outbreaks relies on public sharing of data, and nextstrain has the ability to automatically update sequences from a range of public databases and repositories.
Scientists are justifiably skeptical of losing control of their data, and we try to address these concerns by preventing access to the raw genome sequences, and by clearly indicating the source of each sequence. 
We do, however provide the ability to download phylogenetic trees, metadata and screenshots, and private metadata can be added by users by simply dragging a CSV file onto the browser. 
We believe this strikes a compromise between keeping certain data private and allowing the dissemination of results important to the wider scientific community, and raises the possibility of further collaboration between scientists.
Genomic epidemiology has the potential to inform the public, health organisations and scientists alike, a potential which is realised by sharing of data in real time rather than retrospectively \citep{Croucher:2015go}.

>>>>

Viral pathogens pose an ongoing and ever present danger to human health on a worldwide scale, including epidemics such as the West-African Ebola epidemic and the ongoing Zika epidemic in the Americas, as well as ongoing disease spread such as seasonal influenza.
The rapid evolution of these viruses allows inference of the evolution and epidemiology of the disease from genomic data.
Such analysis are often done in isolation, and may lack the spatial or temporal context in which to best interpret the results, which impedes understanding \citep{pybus2013evolutionary}.
Furthermore, the results of analysis are often not made available to the public or health bodies until publication, which may be too late to effect change in policy or understanding.
We have developed nextstrain to visualise viral outbreaks in as close to real time as possible.
This platform consists of three components: the collection and updating of available sequences from a range of public and private sources, rapid and automated analysis pipelines and an interactive web-application by which to interrogate the results.

The regularly updated nature and rapidity of these analysis is crucial to the monitoring and understanding of pathogen epidemiology and evolution.
Sequencing times and costs are continually dropping, with on-the-ground sequencing being employed during the recent West African Ebola and Zika epidemics \citep{quick2016real,faria2017epidemic}.
This rapidity of sequencing must be complemented with rapid methods by which to analyse and interpret results.

Nextstrain consists of three components:
``Fauna'' is a collection of python scripts to maintain a database representative of all publicly available sequences and related metadata, sourced from public repositories such as NCBI, gisaid and ViPR, as well as github repositories and other sources of genomic data. ``Augur'' is the bioinformatics pipeline, performing subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait reconstruction across the tree, including inference of the most likely transmission events.
This leverages the maximum likelihood phylodynamic analysis TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n=000$) in 0 hours.
These scripts are designed in a manner to be readily co-opted to different pathogens.
Analysed data are then immediately available to interrogate through the visualisation software ``Auspice'', available at \href{www.nextstrain.org}{www.nextstrain.org}.
This approach is similar in concept to Nextflu \citep{neher2015nextflu} however extended and generalised to different viral pathogens.
There is a growing need for surveillance of non-influenza viruses \citep{tang2017global}, and nextstrain is able to be extended to nearly all viral outbreaks with readily accessible genomic data.

@@

For many pathogens, the emergence and spread of gain-of-function variants is a grave concern.
China has experienced seasonal epidemics of influenza A/H7N9 over the past five years, however human to human transmission is thought to be rare, with the majority of human cases being spillover from poultry.
The threat of mutations which facilitate human to human transmission is of extreme concern, as H7N9 has a mortality rate of around $30\%$ \citep{li2014epidemiology}.
\textit{In vitro} experiments have identified three amino acid mutations in the hemagglutinin (HA) gene which confer a switch in specificity for human-type receptors by promoting binding to human trachea epithelial cells \citep{devries2017three}.
Nextstrain allows monitoring of such mutations, such that interventions and control measures may be implemented where appropriate.

\subsection*{A Model for Public Sharing of Data}
Nextstrain presents a single overview of both ongoing viral disease (seasonal \& avian influenza, zika) as well as retrospective analysis (dengue, ebola), all based upon the same underlying bioinformatics architecture.
As such, it is well positioned to respond to both the next outbreak and continued surveillance of current ones.
Analysis of such outbreaks relies on public sharing of data, and nextstrain has the ability to automatically update sequences from a range of public databases and repositories.
Scientists are justifiably skeptical of losing control of their data, and we try to address these concerns by preventing access to the raw genome sequences, and by clearly indicating the source of each sequence.
We do, however provide the ability to download phylogenetic trees, metadata and screenshots, and private metadata can be added by users by simply dragging a CSV file onto the browser.
We believe this strikes a compromise between keeping certain data private and allowing the dissemination of results important to the wider scientific community, and raises the possibility of further collaboration between scientists.
Genomic epidemiology has the potential to inform the public, health organisations and scientists alike, a potential which is realised by sharing of data in real time rather than retrospectively \citep{croucher2015application}.
